---
title: "TreeClim Explore"
output: html_notebook
---

[May 19, 2021] This is a premliminary exploration of the **TreeClim** package. I will focus on the functions. To explore the functions, I will use the examples provided in the manual. 

Functions (in order presented in manual):

1) dcc - response and correlation function analysis
2) dlm - dendro-flavored linear model
3) g_test - Gershunov test for spurious low-frequency modulations
4) seascorr - seasonal (partial) correlation analysis
5) skills - evaluate reconstruction skills using split-calibration
6) traceplot - trace the temporal development of response or correlation functions

Data:



Boosters:

Prepare workspace and load libraries. 
```{r}
rm(list=ls())
library(treeclim)
```

### What is "dcc"?
This function does response and correlation functions. Multiple climate variables can be  input together but it looks like each variable is run individually against the tree-ring chronology. 

```{r}
# to run dcc as a correlation function
dc_corr <- dcc(muc_spruce,muc_clim, method="correlation")

# to run dcc as a response function
dc_resp <- dcc(muc_spruce,muc_clim, method="response")

# fyi: to load the data sets
data("muc_spruce")
data("muc_clim")
```

### What is "dlm"?
This is "dendro-style" SIMPLE (see dlm1 below) or MULTIPLE (see dlm2 below) linear regression, where the predictand is a tree-ring variable and the predictor(s) is/are climate variable(s). 

NOTE: For multiple predictor variables, it does not look like they enter the model in a stepwise fashion.  

Additional notes: 

**Arguments**

*.sum(6:8)* is modified climate variable. in the call below, this modifier acts on the climate data matrix, "rt_prec". 
```{r,message = FALSE}
# Linear model with single predictor
dlm1 <- dlm(rt_spruce,rt_prec,.sum(6:8),param_names = "summer_prec")
dlm1_summ <- summary(dlm1)

# to plot diagnostics
plot(dlm1)

# Linear model with multiple predictors: summer prec and summer temp
dlm2 <- dlm(rt_spruce,list(rt_prec,rt_temp),
            .sum(6:8,"prec")+.mean(6:8,"temp"),
            var_names=c("prec","temp"),
            param_names=c("summer_prec","summer_temp"))

dlm2_summ <- summary(dlm2)
mdl_compare <- anova(dlm1,dlm2)

# to plot diagnostics
plot(dlm2)
```

### What is g-test?
I'm not going to dive too deeply into this at the moment. If desireed, save for future exploration.

```{r eval=FALSE}
dc_cor <- dcc(muc_spruce,muc_clim,3:9,method = "cor",moving = TRUE)
g_test(dc_cor,sb=FALSE)
```
### What is seascorr?

(We know the answer to this!)
```{r}
sc <- seascorr(muc_fake,muc_clim)
sc
plot(sc)
```

### What is skills?

This function evaluates reconstruction skills using split calibration. According to the manual, it can be applied to the following functions: dcc, dlm, or seascorr. 
```{r}
# example from manual
dc <- dcc(muc_fake,muc_clim,.mean(6:9,"temp") + .sum(6:9,"prec"))
sk1 <- skills(dc)
sk1
plot(sk1)
```

```{r}
# from our simple linear regression model that we ran earlier
sk2 <- skills(dlm1)
sk2
plot(sk2)

# skills cannot be run for models with >1 predictor variable
```
### What is traceplot? 
A trace of the temporal development of response or correlation coefficients. 
```{r eval = FALSE}
# note: default method for dcc is "response"
dc_resp <- dcc(muc_spruce,muc_clim,4:9,method = "correlation",dynamic = "evolving")
```

```{r eval = TRUE}
traceplot(dc_resp,c("prec.curr.may","prec.curr.jun","prec.curr.jul"))
```

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
